SILVR: Conditional diffusion model for molecule generation without additional training

### 2023-03-18 paper/manuscript.tex

Typically the objective to achieve is to generate a diverse set of molecules, for a given target protein that are easily synthetically accessible and ideally with a high binding affinity and low toxicity prediction~\cite{some gneric caad}. Different machine learning-based  and other strategies have been used to address these ideas. From Recurrent Neural networks (RNN)~\cite{xu2021novo,zhang2022novo} and conditional variational autoencoders~\cite{ragoza2022generating} to GANs~\cite{skalic2019target}, various approaches have been used to generate molecules for a binding site where in each case the training is conditioned based on the target protein~\cite{xie2022advances}. Some of these models incorporate a ligand score directly~\cite{ragoza2022generating}, while others would require other methods through machine learning~\cite{masters2023deep}, or more traditional docking~\cite{fu2018predictive} and molecular simulation-based affinity prediction methods~\cite{mey2016blinded,mey2020best}. Even with a variety of ways to assess the synthetic accessibility of generated compounds~\cite{coley2018scscore, thakkar2021retrosynthetic} often molecules generated in these method are  not easily sythesiable at best or chemically infeasible in the worst cases. 
Another desirable features for a generative model to have are the ability to generate compounds that are extensions to a known fragment hit or link two existing fragments through different linkers with the desire to improve potency, solubility or other properties of existing hits. Goa et al. have introduced way in which to generate linkers for fragments using reinforcement learning strategies~\cite{guo2023linkinvent}, while Imrie et al.~\cite{imrie2020deep} have used variation atuo encoders on this tasks without the use of protein information. More recently Huang et al.~\cite{huang20223dlinker} and Igashov et al.~\cite{igashov2022equivariant} have tackeled this challenge based on equivariant variational auto-encoder and diffusion based models models respectively. Each of these models were trained for the specific purpose of linker generation.  
Another objective from a medchem perspective is that once a hit has been found is to generate new compounds that improve this hit to become a good lead compound for optimisation.

The work we present here is a framework for conditioning and existing pretrained equivaraint diffusion model (EDM) towards the task of linker generation between fragments and generating similar compounds to existing hits without specific training towards this task.  To achieve this we combine the EDM by Hoogeboom et al.~\cite{hoogeboom2022equivariant} with the iterative latent variable refinement method proposed by Choi et al. for image generation networks~\cite{choi2021ilvr}. This allows the generation of new molecules in the shape of a binding site using information from existing fragments or linking fragments without specific training or knowledge of this task. 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{paper/Figures/Fig1/Schematic.png}
    \caption{Schematic of the equivariant diffusion model with  SILVR indicated for every denoising step.}
    \label{fig:schematic}
\end{figure}

>>>>

Typically the objective to achieve is to generate a diverse set of molecules, for a given target protein that are easily synthetically accessible and ideally with a high binding affinity and low toxicity prediction~\cite{some gneric caad}. Different machine learning-based  and other strategies have been used to address these ideas. From Recurrent Neural networks (RNN)~\cite{xu2021novo,zhang2022novo} and conditional variational autoencoders~\cite{ragoza2022generating} to GANs~\cite{skalic2019target}, various approaches have been used to generate molecules for a binding site where in each case the training is conditioned based on the target protein~\cite{xie2022advances}. Some of these models incorporate a ligand score directly~\cite{ragoza2022generating}, while others would require other methods through machine learning~\cite{masters2023deep}, or more traditional docking~\cite{fu2018predictive} and molecular simulation-based affinity prediction methods~\cite{mey2016blinded,mey2020best}. Even with a variety of ways to assess the synthetic accessibility of generated compounds~\cite{coley2018scscore, thakkar2021retrosynthetic} often molecules generated in these method are  not easily sythesisable at best or chemically infeasible in the worst cases. 
Another desirable features for a generative model to have are the ability to generate compounds that are extensions to a known fragment hit or link two existing fragments through different linkers with the desire to improve potency, solubility or other properties of existing hits. Goa et al. have introduced way in which to generate linkers for fragments using reinforcement learning strategies~\cite{guo2023linkinvent}, while Imrie et al.~\cite{imrie2020deep} have used variation auto encoders on this tasks without the use of protein information. More recently Huang et al.~\cite{huang20223dlinker} and Igashov et al.~\cite{igashov2022equivariant} have tackled this challenge based on equivariant variational auto-encoder and diffusion based models models respectively. Each of these models were trained for the specific purpose of linker generation.  
Another objective from a medchem perspective is that once a hit has been found is to generate new compounds that improve this hit to become a good lead compound for optimisation.

We present, the selective iterative variable refinement (SILVR), a method for conditioning an existing pre-trained equivaraint diffusion model (EDM) towards the task of linker generation between fragments and generating similar compounds to existing hits without specific training towards this task. To achieve this we combine the EDM by Hoogeboom et al.~\cite{hoogeboom2022equivariant} with the iterative latent variable refinement method proposed by Choi et al. for image generation networks~\cite{choi2021ilvr}. This allows the generation of new molecules in the shape of a binding site using information from existing fragments or linking fragments without specific training or knowledge of this task.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{paper/Figures/Fig1/Schematic.png}
    \caption{Schematic of the equivariant diffusion model with  SILVR indicated for every denoising step.}
    \label{fig:schematic}
\end{figure}

Denoising probabilistic diffusion models can be separated into two parts, the diffusion part and the denoising part, as shown in the schematic in Figure~\ref{fig:schematic} with the ideas originating in image generation machine learning problems. A neural network is trained to learn the second part of the process, denoising a Gaussian distribution until an image--or in this case--a molecule is generated. The idea we propose here is to effectively introduce information of a reference molecule at a given rate to the denoising process of a pre-trained model. This, similar to the concept of inpaining~\cite{qin2021image, xie2012image, squires2023artefact}, guides the denoising process at each step towards the reference. In this paper we show, that we can generate new compounds that are similar to a given reference and even link fragments using multiple references as input. The proposed method is generalizeable to any protein ligand systems with known fragment hits. For illustration purposes we use the original 23 X-ray fragment hits from the Covid-Moonshot dataset~\cite{consortium2021open,consortium2023open}. In the following, we give an overview of how equivariant difusion models work and introduce how our method SILVR fits into the framework of an existing pre-trained EDM. We then show, how the SILVR rate, $r_{\mathrm{S}}$ as modelling parameter can modulate how much of the reference is incorporated in the sample generation and how the original EDM is recovered at  $r_{\mathrm{S}}=0$. As part of this we illustrate how fragments can be linked using dummy atoms and how newly generated molecules can fit into the existing binding site according to shape similarity. Some of the limitations that arise from the proposed methods are connected to the chosen pre-trained EDM with which we conclude our discussions.
